Rstudio in Data Science and Paper Writing

### 2024-11-02 presentation/RStudio_paper_writing_course_presentation_part1.qmd
talk update: split some slides, optimise layout, make presentation and course exercise same

---
title: "Rstudio in Data Science and Paper Writing"
author: "Gáspár Jékely"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: false
    code-fold: false
    code-summary: "Code"
    chalkboard: 
      buttons: false
    preview-links: auto
scrollable: true
---

@@

<img src=https://media1.tenor.com/images/f6362876996697b6a6f554b2ac3d3013/tenor.gif?itemid=10488408 width=200>

@@

- do you have a GitHub account? 
- navigate to: https://github.com/JekelyLab/RStudio_paper_writing_course
- fork the repository to your GitHub (if you have an account already)
- go to RStudio -> New Project -> Git -> your repo's address (or my address)
- open /analysis/scripts/Course_exercises.R

@@

@@

-   Whenever you refer to a file with a relative path it will look for it in your wd.

@@

@@

* We will use https://github.com/JekelyLab/RStudio_paper_writing_course

* Go to github and clone the project to your account ("Use this template")

* Go to RStudio -> new project -> version control -> git -> git URL of your new project

* select local dir and download the project with all directories to your computer

* let's check the project

* you can rename the .Rproject file to 'my favourite file name'

@@

## rio: A Swiss-Army Knife for Data I/O

@@

- once your data have been tidied up, plotting works like a charm
- use ggplot2 and a tibble as input
- ggplot2 is part of tidyverse and loads when you load tidyverse
- very versatile and extendable data visualisation package
- as input, you need a tibble (or a data.frame)
- supply data, define aesthetic mapping, add layers, define axes, plot theme etc.
- use the same theme (font size, line width etc.) across all figures in a project/paper/thesis
- statistical tests
- saving plots and source data

@@

@@

```{python}
import matplotlib
print(matplotlib.__version__)
```

@@

@@

img1 <- readPNG("../analysis/pictures/plot_Jose1a.png")
img2 <- readPNG("../analysis/pictures/plot_Jose1b.png")

panel_JoseA <- ggdraw() + draw_image(img1)
panel_JoseB <- ggdraw() + draw_image(img2)

@@

#save figure as png and pdf
ggsave("../manuscript/figures/Figure_Jose.png", limitsize = FALSE, 
       units = c("px"), Figure_Jose, width = 4000, height = 2000,
       bg = "white")

ggsave("../manuscript/figures/Figure_Jose.pdf", limitsize = FALSE, 
       units = c("px"), Figure_Jose, width = 3000, height = 1600)

@@

#add text labels
panel_INNOS <- ggdraw() + 
  draw_image(img_INNOS) +
  draw_label("INNOS", x = 0.3, y = 0.99, size = 10) +
  draw_label("NS plexus", x = 0.485, y = 0.59, size = 8) +
  draw_label("outgoing", x = 0.9, y = 0.45, size = 10, color='#E69F00') +
  draw_label("incoming", x = 0.89, y = 0.5, size = 10, color='#0072B2') +
  draw_label("D", x = 0.95, y = 0.93, size = 6) +
  draw_label("V", x = 0.95, y = 0.77, size = 6) +
  draw_label("*", x = 0.5, y = 0.29, color='black',size = 18,fontface='plain') +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = arrow, 
               arrow = arrow(ends = "both", type = "closed", length = unit(0.1,"cm")),
               lineend = "butt",
               linejoin = "mitre",
               arrow.fill = "black", size = 0.2)

#define layout
layout <- "AB"

@@

#save figure as pdf
ggsave("../manuscript/figures/Figure_IHC.pdf", limitsize = FALSE, 
       units = c("px"), Figure_INNOS, width = 3000, height = 1000)

image_read("../manuscript/figures/Figure_IHC.png")
#img

@@

@@

### Excercises

@@

#introduce gap in layout
layout <- "
AAAABBBBCCCC
AAAABBBBDDDD
############
EEEFFFGGGHHH
EEEFFFGGGHHH
"

@@

#save figure as png
ggsave(
  "../manuscript/figures/Figure_complex.png",
  units = c("px"), Figure_complex, 
  width = 2600, height = 1700, bg = "white"
  )

image_read("../manuscript/figures/Figure_complex.png")
```

@@

-   -\> Open manuscript/Manuscript.qmd\]

@@

---
title: "New paper template"
subtitle: |
  \raggedright  Author 1^1,2,\*^ , Author 2^1,2^, Author 3^3^, Author 4E^4^ \newline<br>^1^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^2^Affiliation 2 \newline<br>^3^Affiliation 3 \newline<br>^4^Affiliation 4\newline<br>  ^\*^Correspondence: xy@cos.uni-heidelberg.de
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

@@

-   In Visual mode, go to 'Insert' -\> 'Citation'
-   search for reference on pubmed etc.
-   'Insert' will add the ref to the bibliograpy.bib file
-   in 'Source' mode, you can directly insert the reference with [@ref_id] (preferred, faster method)
-   add new refs to your .bib file with the command-line command:

@@

::: callout-warning
-   need to add references.bib to the YAML header as bibliography: references.bib
-   also define bibliography style in the YAML with csl: elife.csl #add .csl according to journal
:::

>>>>

---
title: "Rstudio in Data Science and Paper Writing"
author: "Gáspár Jékely <br> Centre for Organismal Studies, Heidelberg University"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: false
    code-fold: false
    code-summary: "Code"
    chalkboard: 
      buttons: false
    preview-links: auto
scrollable: true
---

@@

<img src=https://media1.tenor.com/images/f6362876996697b6a6f554b2ac3d3013/tenor.gif?itemid=10488408 width=160>

@@

- do you have a GitHub account? 
- navigate to: https://github.com/JekelyLab/RStudio_paper_writing_course
- fork the repository to your GitHub (if you have an account already)
- go to RStudio -> New Project -> Git -> your repo's address (or my address)
- select local dir and download the project with all directories to your computer
- let's check the project
- you can rename the .Rproject file to 'my favourite file name'
- open /analysis/scripts/Course_exercises.R

@@

## Use R projects (.Rproj)

@@

-   Whenever you refer to a file with a relative path it will look for it in your working dir.

- in case you used setwd() in a script (you should not) you can find your Rproject directory with here()

```{r, eval=TRUE, echo=TRUE}
here::here()

#test, but you should not do this
setwd("~")
getwd()

#get back to Rproj dir
setwd(here::here())
getwd()
# all good again
```

## Use R projects (.Rproj)

- list files in your working dir

@@

<br><br><br>

@@

* Check the folder structure in the downloaded project template

@@

## File import and saving -- rio: A Swiss-Army Knife for Data I/O

@@

- once your data have been tidied up, plotting is a lot easier
- use ggplot2 and a tibble as input
- ggplot2 is part of tidyverse and loads when you load tidyverse
- very versatile and extendable data visualisation package
- as input, you need a tibble (or a data.frame)
- supply data, define aesthetic mapping, add layers, define axes, plot theme etc.
- use the same theme (font size, line width etc.) across all figures in a project/paper/thesis
- statistical tests
- saving plots and source data

@@

## Faceting

- Faceting means spreading the data across different panels

- Need to define 'facets', a set of variables or expressions quoted by vars() defining faceting groups

```{r eval = TRUE, echo = TRUE, fig.height = 4}
iris %>%  
  ggplot(aes(
    x = Petal.Length, y = Sepal.Width, 
    color = Species, size = Sepal.Width, shape = Species) 
    ) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(Species))
```

@@

```{python, echo=TRUE}
import matplotlib
print(matplotlib.__version__)
```

@@

# Add significance values of stats

plot_Ashwini_ct +
  stat_compare_means(
    aes(label = after_stat(p.format)),
    method = "wilcox.test",
    #ref.group = "Foxg1"
    comparisons = list( c("GAPDH", "Nanog"), c("GAPDH", "oct4"), c("GAPDH", "pax6") )
    )
```

@@

img1 <- readPNG("../analysis/pictures/plot_Jose1a.png")
img2 <- readPNG("../analysis/pictures/plot_Jose1b.png")
panel_JoseA <- ggdraw() + draw_image(img1)
panel_JoseB <- ggdraw() + draw_image(img2)

@@

#save figure as png and pdf
ggsave(
  "../manuscript/figures/Figure_Jose.png", limitsize = FALSE, 
  units = c("px"), Figure_Jose, width = 4000, height = 2000, bg = "white"
  )
ggsave(
  "../manuscript/figures/Figure_Jose.pdf", limitsize = FALSE, 
  units = c("px"), Figure_Jose, width = 3000, height = 1600
  )
```

```{r}
#| echo=FALSE, eval=TRUE

@@

#add text labels
panel_INNOS <- ggdraw() + 
  draw_image(img_INNOS) +
  draw_label("INNOS", x = 0.3, y = 0.99, size = 10) +
  draw_label("NS plexus", x = 0.485, y = 0.59, size = 8) +
  draw_label("outgoing", x = 0.9, y = 0.45, size = 10, color='#E69F00') +
  draw_label("incoming", x = 0.89, y = 0.5, size = 10, color='#0072B2') +
  draw_label("D", x = 0.95, y = 0.93, size = 6) +
  draw_label("V", x = 0.95, y = 0.77, size = 6) +
  draw_label("*", x = 0.5, y = 0.29, color='black',size = 18,fontface='plain') +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = arrow, 
               arrow = arrow(ends = "both", type = "closed", length = unit(0.1,"cm")),
               lineend = "butt",
               linejoin = "mitre",
               arrow.fill = "black", size = 0.2)
```

```{r}
#| echo=TRUE, eval=TRUE
#define layout
layout <- "AB"

@@

#save figure as pdf
ggsave(
  "../manuscript/figures/Figure_IHC.pdf", limitsize = FALSE, 
  units = c("px"), Figure_INNOS, width = 3000, height = 1000
  )
```

```{r}
#| echo=FALSE, eval=TRUE

image_read("../manuscript/figures/Figure_IHC.png")
```

@@

## Reading tif files, parsing metadata like image scale

```{r}
#| echo=TRUE, eval=TRUE
#| 
img_tif <- magick::image_read("../analysis/pictures/INNOS_synapses.tif")
img_scale <- image_info(img_tif)$width/as.numeric(str_split(image_info(img_tif)$density, "x")[[1]][1])
img_scale #not precise

# alternatively use ijtiff
library(ijtiff)
img_tif_ij <- ijtiff::read_tags("../analysis/pictures/INNOS_synapses.tif")
img_scale <- img_tif_ij$frame1$width/img_tif_ij$frame1$x_resolution
img_scale

panel_INNOS_tif <- ggdraw() + draw_image(img_tif)

Figure_INNOS_with_tif <- plot_syn_ann + panel_INNOS_tif +
  plot_layout(design = layout, widths = c(2, 1)) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 12, face='plain'))

#save figure as png
ggsave(
  "../manuscript/figures/Figure_INNOS_with_tif.png", limitsize = FALSE,
  units = c("px"), Figure_INNOS_with_tif, 
  width = 3000, height = 1000, bg = "white"
  )
```

```{r}
#| echo=FALSE, eval=TRUE

image_read("../manuscript/figures/Figure_INNOS_with_tif.png")
```

@@

### Exercises

@@

#introduce gap in layout
layout <- "
AAAABBBBCCCC
AAAABBBBDDDD
############
EEEFFFGGGHHH
EEEFFFGGGHHH
"
```

@@

#save figure as png
ggsave(
  "../manuscript/figures/Figure_complex.png",
  units = c("px"), Figure_complex, 
  width = 2600, height = 1700, bg = "white"
  )
```

```{r}
#| echo=FALSE, eval=TRUE
image_read("../manuscript/figures/Figure_complex.png")
```

@@

-   -\> Open manuscript/Manuscript.qmd

@@

---
title: "Paper or report template"
subtitle: |
  \raggedright  Author 1^1,2,\*^ , Author 2^1,2^, Author 3^3^, Author 4E^4^ \newline<br>^1^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^2^Affiliation 2 \newline<br>^3^Affiliation 3 \newline<br>^4^Affiliation 4\newline<br>  ^\*^Correspondence: xy@cos.uni-heidelberg.de
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: 
    embed-resources: true
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

## Simple syntax

- [This is]{.underline} 
- A **Quarto** *document.* 
- Quarto uses a simple ^Markdown^ formatting syntax for authoring HTML, PDF, and MS Word or .odt documents. - Here is some **text in bold** *italic* [underline]{.underline} and ^superscript^.

@@

-   In Visual mode, go to 'Insert' -\> 'Citation'
-   search for reference on pubmed etc.
-   'Insert' will add the ref to themanuscript/ bibliograpy.bib file (needs to be defined in the Yaml header)
-   in our doc, references will be formatted in the eLife style (defined in the Yaml header under: csl: elife.csl)
-   The elife.csl file is saved in the manuscript folder.
-   For a different format, download the respective .csl file (e.g., from the Zotero style repository <https://www.zotero.org/styles>)
-   in 'Source' mode, you can directly insert the reference with [@ref_id] (preferred, faster method)

-   you can also add new refs to your .bib file with the command-line command:

@@

::: callout-warning
-   need to add references.bib to the YAML header as bibliography: references.bib
-   also define bibliography style in the YAML with csl: elife.csl #add .csl according to journal
- if you use the command line, use >> and not > otherwise you will overwrite your file!
:::

## Inserting figures

-   we can now insert the figures we generated previously and saved under /manuscript/figures or /manuscript/figure_supplements

-   Figure title and legend can also be edited, as well as width

```{r, out.width = "100%", fig.cap="**Figure 1. A figure** (A) A nice picture. (B) legend.  (C) <br> (D) " }
#| echo: true

knitr::include_graphics("../manuscript/figures/Figure1.png")
```

## Adding equations

- Equations can also be inserted, Insert -\> Display Math:

$$
\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
$$
## Sourcing code and working with variable

```{r}
#| eval: true
source("../analysis/scripts/statistics_for_paper.R")
```

-   We can use the 'r variable' syntax to add variables calculated in the sourced code directly to the text
-   This way if the underlying dataset changes, we just reprocess a calculation and insert the result directly into our manuscript
- the script is sourced and it runs but the output is not included in the knitted output. But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is `r max_PRC` (now this number comes from our sourced script).
